Association between trauma severity models and opportunities for improvement:

### 2022-09-22 studyprogressreport.Rmd
updated according to martins comments, done

Introduction
============
Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic<!-- I typically define trauma as the clinical entity composed of physical injury and the body's associated response. I would suggest that you stick to trauma and don't use physical injury and trauma interchangeably -->. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

## Trauma severity scores
Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury. An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma. This question has spurred the development of multiple trauma severity scores (TSS)<!-- Avoid non-standard abbreviations, as it makes your text harder to read. For example, I don't think spelling out trauma severity scores is such a big deal, especially considering that you'll be dealing with so many score names that are abbreviations themselves -->. TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality. TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview]. Today there are hundreds of TSS models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).

Trauma scoring has become a cornerstone of modern trauma QI programs [@historytss]. One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines]. TSS also allow clinics to compare their outcomes against an established benchmark [@whoguidelines]. However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from. Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality]. This weakness is one reason why models continue to be developed and improved on. Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.

<!-- I know that a lot of the following details about TRISS/NORMIT are probably not really relevant to my aim, it's mainly here to fill out on length -->

@@

## Trauma severity scores as quality indicators

Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring. However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.

If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion. Conversely, if TSS models are not associated with OFI, using TSS as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.

There are far more TSS models than the two described above. These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is where this institution is located. It seems reasonable to begin by investigating these two models. The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.

<!-- Note that you mix trauma severity scores, trauma severity models, prediction models, etc. Stick to one of these concepts -->

@@

<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? --> <!-- That's fine -->

@@

This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysed to estimate the association between the common TSS models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> <!-- Don't include numbers here, those belongs in Reults -->

@@

### Exposures
The primary exposures of interest are trauma severity scores according to the TRISS and NORMIT models. Both of these models estimate the Ps. This is expressed a probability and is a continuous variable. The variables integrated into these models are taken from arrival at hospital.

@@

<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. Do I need to have the formulas? --> <!-- Yes, you need the formulas, but not for this report -->

@@

The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age. The scores will be calculated according to the formulas given by the respective developers of each score.

<!-- BTW this is here because, according to me and what I see in the database, it seems like I'll have to calculate NORMIT/TRISS for everyone based on parameters in the database -->

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. In cases where data necessary for analysis is missing, i.e. the variables referred to above, the participant will be excluded from the study. Logistic regression will be used to estimate the predictive value of each TSS in regards to OFI. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant.

@@

It is possible that scores other than NORMIT and TRISS will be analysed if there is time and it is feasible based on the data I have access to. I will decide this within the first week or two of starting my data analysis code.

>>>>

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response <!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

## Trauma severity models
Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury. An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma. This question has spurred the development of multiple trauma severity models. Trauma severity models attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality. Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview]. Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).

Trauma severity models have become a cornerstone of modern trauma QI programs [@historytss]. One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines]. Trauma severity models also allow clinics to compare their outcomes against an established benchmark [@whoguidelines]. However, an inherent weakness of any trauma severity model is that estimates may be unreliable when applied to a population different from the one the model was developed from. Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality]. This weakness is one reason why models continue to be developed and improved on. Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.

@@

## Trauma severity models as quality indicators

Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring. However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether trauma severity models have better performance when taking into account the entire population of trauma patients.

If it could be found that an existing trauma severity model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion. Conversely, if trauma severity models are not associated with OFI, using trauma severity models as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.

There are far more trauma severity models than the two described above. These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is where this institution is located. It seems reasonable to begin by investigating these two models. The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.

@@

@@

This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria.

@@

### Exposures
The primary exposures of interest are the trauma severity models TRISS and NORMIT. Both of these models estimate the Ps. This is expressed a probability and is a continuous variable. The variables integrated into these models are taken from arrival at hospital.

@@

@@

NORMIT and TRISS will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age. The scores will be calculated according to the formulas given by the respective developers of each model.

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. In cases where data necessary for analysis is missing, i.e. the variables referred to above, the participant will be excluded from the study. Logistic regression will be used to determine association between the estimates of the trauma severity models and OFI. To assess how well the models predict OFI measures of discrimination and calibration will be used. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant.

@@

It is possible that models other than NORMIT and TRISS will be analysed if there is time and it is feasible based on the data I have access to. I will decide this within the first week or two of starting my data analysis code.
